Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models

### 2015-02-03 revision_v2/Paper_v3.tex
second rebuttal - new paragraph

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort&compress,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

%\usepackage{geometry}
\usepackage{fullpage}

%\bibliographystyle{plain}
\bibliographystyle{plos2009}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}	
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

@@

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective models of metabolism}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\textit{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated. Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information. Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data 
starting from an unbiased collection of possible allosteric structures using particle swarm optimization. However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure. While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.

{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}

\pagebreak

\setcounter{page}{1}

\linenumbers

\section*{Introduction}
Mathematical modeling has long contributed to our understanding of metabolism. 
Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}. 
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously. 
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

\clearpage

\section*{Results}

\subsection*{Formulation and properties of effective cell-free metabolic models.}
We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.

Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig. \ref{fig-control-schematic}).  
This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}. 
In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites, however in general this can be a complex function of both modeled and unmodeled factors. 
When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time. 
Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.

@@

@@

\subsection*{Estimating parameters and effective allosteric regulatory structures.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
We explored this linkage by estimating model parameters from synthetic data using both network structures. 
We generated synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
network A. This data set is similar to published cell-free studies, both in terms of network coverage and sampling frequency \citep{Jewett:2008aa}. 
We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B). 
However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.

@@

@@

\clearpage

>>>>

%  LaTeX support: latex@mdpi.com
%  In case you need support, please attach any log files that you could have, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

%=================================================================

% LaTeX Class File and Rendering Mode (choose one)
% You will need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.

%=================================================================

\documentclass[processes,article,accept,moreauthors,pdftex,12pt,a4paper]{mdpi}

\usepackage{url}  % Formatting web addresses  
%\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
%\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{soul}

%\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
%\usepackage{setspace}
%\usepackage{longtable}
%\usepackage{colortbl}
%\usepackage{palatino,lettrine}
\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
%\usepackage[square,sort&compress,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
%\usepackage{lineno}
%
%\urlstyle{rm}

\setcounter{page}{1}
\lastpage{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2014}
\history{Received: xx / Accepted: xx / \\
Published: xx}
%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
%\doublespace

%\usepackage{geometry}
%\usepackage{fullpage}

%\bibliographystyle{plain}
%\bibliographystyle{plos2009}

%\makeatletter
%\renewcommand\subsection{\@startsection
%	{subsection}{2}{0mm}
%	{-0.05in}
%	{-0.5\baselineskip}
%	{\normalfont\normalsize\bfseries}}
%\renewcommand\subsubsection{\@startsection
%	{subsubsection}{2}{0mm}
%	{-0.05in}
%	{-0.5\baselineskip}
%	{\normalfont\normalsize\itshape}}
%\renewcommand\section{\@startsection
%	{subsection}{2}{0mm}
%	{-0.2in}
%	{0.05\baselineskip}
%	{\normalfont\large\bfseries}}	
%\renewcommand\paragraph{\@startsection
%	{paragraph}{2}{0mm}
%	{-0.05in}
%	{-0.5\baselineskip}
%	{\normalfont\normalsize\itshape}}
%\makeatother

@@

% Single space'd bib -
%\setlength\bibsep{0pt}

%\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% Change the number format in the ref list -
%\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Figure
%\renewcommand{\figurename}{Figure}
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, administrativesciences, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, appliedsciences, arts, atmosphere, atoms, axioms, batteries, behavioralsciences, bioengineering, biology, biomedicines, biomolecules, biosensors, brainsciences, buildings, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, cosmetics, crystals, dentistryjournal, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environmentalsciences, environments, epigenomes, fibers, foods, forests, futureinternet, galaxies, games, gels, genes, geosciences, healthcare, horticulturae, humanities, hydrology, informatics, information, inorganics, insects, ijerph, ijfs, ijms, ijgi, jcdd, jcm, jdb, jfb, jof, joi, jlpea, jmse, jpm, jrfm, jsan, land, laws, life, lubricants, machines, marinedrugs, materials, mathematics, medicalsciences, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, nanomaterials, ncrna, nutrients, pathogens, pharmaceuticals, pharmaceutics, pharmacy, photonics, plants, polymers, processes, proteomes, publications, religions, remotesensing, resources, risks, robotics, safety, sensors, sinusitis, socialsciences, societies, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, universe, vaccines, veterinarysciences, viruses, water
%---------
% article
%---------
% The default type of manuscript is article, but could be replaced by using one of the class options: 
% article, review, communication, commentary, bookreview, correction, addendum, editorial, changes, supfile, casereport, comment, conceptpaper, conferencereport, meetingreport, discussion, essay, letter, newbookreceived, opinion, projectreport, reply, retraction, shortnote, technicalnote, creative
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g. the logo of the journal will get visible), the headings, and the copyright information. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
% Please insert a blank line is before and after all equation and eqnarray environments to ensure proper line numbering when option submit is chosen
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option "pdftex" is for use with pdfLaTeX only. If eps figure are used, use the optioin "dvipdfm", with LaTeX and dvi2pdf only.

%=================================================================

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================

% Add packages and commands to include here
% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are loaded by the MDPI class.
%\usepackage{graphicx}
%\usepackage{subfigure,psfig}

%=================================================================
%% Please use the following mathematics environments:
%\theoremstyle{mdpi}
%\newcounter{thm}
%\setcounter{thm}{0}
%\newcounter{ex}
%\setcounter{ex}{0}
%\newcounter{re}
%\setcounter{re}{0}
%\newtheorem{Theorem}[thm]{Theorem}
%\newtheorem{Lemma}[thm]{Lemma}
%\newtheorem{Characterization}[thm]{Characterization}
%\newtheorem{Proposition}[thm]{Proposition}
%\newtheorem{Property}[thm]{Property}
%\newtheorem{Problem}[thm]{Problem}
%\newtheorem{Example}[ex]{Example}
%\newtheorem{Remark}[re]{Remark}
%\newtheorem{Corollary}[thm]{Corollary}
%\newtheorem{Definition}[thm]{Definition}
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================

% Full title of the paper (Capitalized)
\Title{Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}

% Authors (Add full first names)
\Author{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner {*}}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853, USA\\ \vspace{12pt}

Running Title:~Effective models of metabolism}

% Contact information of the corresponding author (Add [2] after \corres if there are more than one corresponding author.)
\corres{E-Mail: jdv27@cornell.edu; \mbox{Tel.: +607-255-4258;} Fax: +607-255-9166.}

\abstract{
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated. Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information. Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data 
starting from an unbiased collection of possible allosteric structures using particle swarm optimization. However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure. While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.}

% Keywords: add 3 to 10 keywords
\keyword{Cell-free metabolism, Mathematical modeling}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{}
%\MSC{}
%\JEL{}

% If this is an expanded version of a conference paper, please cite it here: enter the full citation of your conference paper, and add $^\dagger$ in the end of the title of this article.
%\conference{}

\begin{document}

\section{Introduction}

Mathematical modeling has long contributed to our understanding of metabolism. Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}. 
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously. 
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

\newpage

\section{Results}\vspace{-12pt}

\subsection{Formulation and properties of effective cell-free metabolic models.}

We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./figs/Figure-1-Networks.pdf}
\caption{Proof-of-concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation.}\label{fig-networks}
\end{figure}

Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig. \ref{fig-control-schematic}).  
This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}. 
In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites, however in general this can be a complex function of both modeled and unmodeled factors. 
When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time. 
Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.  
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
\caption{Schematic of rule-based allosteric enzyme activity control laws. 
Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
For a fixed control gain parameter $\kappa_{control}$, the reaction rate approached a step for increasing cooperativity control parameter $\eta$. 
For activation simulations $\kappa_{control}=$0.05 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.
\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
For a fixed control gain parameter $\kappa_{control}$, the reaction rate approximated non-competitive inhibition for increasing cooperativity control parameter $\eta$. 
For the inhibition simulations $\kappa_{control}=$1.5 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.}\label{fig-kinetics-simulations}
\end{figure}

@@

\subsection{Estimating parameters and effective allosteric regulatory structures.}

A critical challenge for any dynamic model is the estimation of kinetic parameters. 
For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
We explored this linkage by estimating model parameters from synthetic data using both network structures. 
We generated synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
network A. This data set is similar to published cell-free studies, both in terms of network coverage and sampling frequency \citep{Jewett:2008aa}. 
We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B). 
However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.

\begin{figure}[H]
\centering
\includegraphics[width=0.450\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
\caption{ON/OFF control simulations for Network A and Network B for an ensemble of 100 kinetic parameter sets versus time. 
For each case, simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. 
Control parameters were fixed during the ensemble calculations.
\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
no inhibition case (Control-OFF).
\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).}\label{fig-onoff-simulations}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
enzymatic and control connectivity starting from randomized initial parameters.
\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
the ensemble of models did not describe the synthetic data. The synthetic data plotted here was unperturbed by noise. 
However, we assumed a constant coefficient of variation of 10\% for the synthetic data during parameter estimation.}\label{fig-parameter-fit}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.47\textwidth]{./figs/Figure-7-ControlSearch.pdf}
\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
A population of 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
Simulation error was minimized for a synthetic data set 
($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
\textbf{A}: Simulation error versus parameter set angle for 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
\textbf{B}: Simulation error versus parameter set angle for 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
\textbf{C}: Simulation error versus parameter set angle for 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
a large incorrect bias. }\label{fig-control-search}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
The particle error, the control variables governing $E_{1},E_{3}$ and $E_{6}$ activity ($v_{1},v_{3}$ and $v_{3}$) and the scaled metabolic flux were calculated for the positively (top), 
uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 
Blue denotes a low value, while red denotes a high value for the respective quantity being plotted.
The particles from each swarm were sorted based upon simulation error (low to high error). 
\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
This is especially apparent in the negatively biased particle swarm. }\label{fig-flux-pattern}
\end{figure}

@@

\section{Discussion}
